# Introduction

This vignette supports workshops on road safety analysis with R.
It was initially developed for a 2 day [Introduction to R for Road Safety](https://www.racfoundation.org/introduction-to-r-for-road-safety) course funded by the [RAC Foundation](https://www.racfoundation.org/), a charity that supports road safety research.

The course is based on open access road crash data from the UK, which is provided in an accessible form to R users through the R package **stats19** [@lovelace_stats19_2019].
However, the content should be of use for anyone working with road crash data that has (at a minimum):

- A timestamp
- A location (or address that can be geocoded)
- Attribute data, such as severity of crash, mode of vehicles involved etc.

You should type, run and ensure you understand each line of code in this document.

The '[issue tracker](https://github.com/ropensci/stats19/issues)' associated with that repo is a good place to ask questions for the community. 

## Why R for road safety research?

Todo.

## Prerequisites

If you are not experienced with R, it is strongly advised that you read-up on and more importantly **test out** R and RStudio before attempting analyse road crash data with R.
See the `stats19-training-setup` vignette at https://docs.ropensci.org/stats19/articles/stats19-training-setup.html for guidance on getting started with R, RStudio and installing R packages.

The completing the course requires that the following packages, which can be installed with `install.packages()`, can be loaded as follows:

```{r, message=FALSE, warning=FALSE, eval=FALSE}
library(pct)      # access travel data from DfT-funded PCT project 
library(sf)       # spatial vector data classes
library(stats19)  # get stats19 data
library(stplanr)  # transport planning tools
library(tidyverse)# packages for 'data science'
library(tmap)     # interactive maps
```

You should type, run and ensure you understand each line of code in this document.


```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  out.width = "50%"
)
```

```{r pkgs, warning=FALSE, echo=FALSE}
pkgs = c(
  "sf",          # spatial data package
  "stats19",     # downloads and formats open stats19 crash data
  "dplyr",       # a package data manipulation, part of the tidyverse
  "tmap"         # for making maps
)
```

```{r cite, echo=FALSE}
knitr::write_bib(x = pkgs, "packages.bib")
```

```{r, eval=FALSE, echo=FALSE}
remotes::install_cran(pkgs)
# remotes::install_github("ITSLeeds/pct")
```

The workshop is based on a practical handout that can be found [here](https://github.com/ropensci/stats19/releases/download/1.1.0/stats-19-exercises.pdf).
