# Introduction

This book teaches reproducible road safety analysis with R.
It was initially developed for a 2 day [Introduction to R for Road Safety](https://www.racfoundation.org/introduction-to-r-for-road-safety) course funded by the [RAC Foundation](https://www.racfoundation.org/), a charity that supports road safety research, which funded the development of this free and open resource.

The content is based on open access road crash data from the UK, which is provided by the R package **stats19** [@lovelace_stats19_2019], although the content should help anyone working with road crash data that has (at a minimum):

- A timestamp
- A location (or address that can be geocoded)
- Attribute data, such as severity of crash, and type of vehicles involved

The book is practical, meaning that you should reproduce the examples that are provided throughout.
As with many practical skills you learn data science by *doing* data science.

Before getting stuck in with the practical content, which begins in Chapter \@ref(r-basics), the rest of this chapter introduces the concept reproducibility and its importance for evidence-based policies, explains why R is an excellent tool for reproducible data science, and outlines how to install R on your computer or access it through remote servers in the 'cloud'.
Finally, Section \@ref(overview) explains the structure of the book and providers pointers on the contents of the chapters and how they should be used for maximum benefit, depending on your level of experience and aims.

## Reproducibility



## Why R for road safety research?

Todo.

## Prerequisites

If you are not experienced with R, it is strongly advised that you read-up on and more importantly **test out** R and RStudio before attempting analyse road crash data with R.
See the `stats19-training-setup` vignette at https://docs.ropensci.org/stats19/articles/stats19-training-setup.html for guidance on getting started with R, RStudio and installing R packages.

The completing the course requires that the following packages, which can be installed with `install.packages()`, can be loaded as follows:

```{r, message=FALSE, warning=FALSE, eval=FALSE}
library(pct)      # access travel data from DfT-funded PCT project 
library(sf)       # spatial vector data classes
library(stats19)  # get stats19 data
library(stplanr)  # transport planning tools
library(tidyverse)# packages for 'data science'
library(tmap)     # interactive maps
```

You should type, run and ensure you understand each line of code in this document.


```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  out.width = "50%"
)
```


```{r pkgs, warning=FALSE, echo=FALSE}
pkgs = c(
  "sf",          # spatial data package
  "stats19",     # downloads and formats open stats19 crash data
  "dplyr",       # a package data manipulation, part of the tidyverse
  "tmap"         # for making maps
)
```

```{r cite, echo=FALSE}
knitr::write_bib(x = pkgs, "packages.bib")
```

```{r, eval=FALSE, echo=FALSE}
remotes::install_cran(pkgs)
# remotes::install_github("ITSLeeds/pct")
```

## Installing R and RStudio


## R in the cloud

## Overview

Todo: add brief introduction to the rest of the book.
