# Temporal data {#time}

Time is ubiquitous in road safety data: collisions and road safety implementations always happened at some point in time.
Before demonstrating how to handle time series data in base R and with `hms` and `lubridate` packages, in subsequent sections, we will show how you can analyse the temporal dimensions of the real world `crashes_2019` object we created in Chapter \@ref(pkgs).
The aim is to get you up-to-speed with how data analysis with time data 'feels' before learning the details in subsequent sections.
If you're the kind of person who likes to know the details first, feel free to skip the next section and return to it after having read the other sections.

## Temporal analysis of crash data

To get a feel for temporal data analysis in R, let's start by reading-in crash data for 2019 with the `stats19` package by typing the following into the Source Editor and running the code line-by-line, as taught in Chapter \@ref(rstudio):

```{r, message=FALSE}
library(stats19)
crashes_2019 = get_stats19(2019)
```

Note that, unlike the longer `crashes_2019 = get_stats19(year = 2019, type = "accidents")` used in Chapter \@ref(pkgs), we did not use *named arguments* in this code chunk.
Instead of `year = 2019`, we simply typed 2019.
That is possible because R functions can be specified by name matching or order: the first argument of `get_stats()` is `year` so the function is expecting a year value.
Also, although we didn't explicitly specify the `accidents` table, `type = "accidents"` is the default value, so `type` only needs to be specified when importing casualty and vehicle datasets.

With that educational aside out of the way, let's take a look at the time variables that are actually in out newly read-in dataset:

```{r}
library(tidyverse)
crashes_2019 %>% 
  select(matches("time|date")) %>% 
  names()
```

Building on the previous chapter and a bit of guesswork, it should be clear what just happened: 
we selected variables that *match* (with the `matches()` function) the character strings `"time"` or (as indicated by the `|` vertical pipe symbol) `"date"` and returned the matching variable names.
This shows that the `stats19` package gives you not one, not two, but three temporal variables.

Exercises:

1. Print the first 6 and then the first 10 elements of each of the three temporal variables in `crashes_2019`
1. What is the class of each variable (technically, of each vector)?
1. Bonus: Extract the weekday from the variable called `date`.
1. Bonus: How many crashes happened on Monday?

```{r, eval=FALSE}
library(stats19)
crashes_2017 = stats19::get_stats19(year = 2017, type = "ac")
crashes_2017
```

<!-- **Advanced challenge:** calculate how many crashes occurred for each day of the week. Then plot it with ggplot2. Repeat the same exercises extracting the hour of the car accident from the variable called time. How would you combine the two informations in a single plot?  -->

```{r, eval=FALSE, echo=FALSE}
# solutions
crashes %>% filter(casualty_hour >= 12)
crashes %>% filter(casualty_hour > 15 & casualty_hour < 19)

crashes_2017 %>% 
  mutate(my_weekdays = weekdays(date)) %>%
  filter(my_weekdays == "Monday") %>% 
  nrow()
crashes_2017 %>% 
  mutate(my_weekdays = weekdays(date)) %>%
  filter(my_weekdays == "Friday") %>% 
  nrow()

crashes_2017 %>% 
  mutate(my_weekdays = weekdays(date)) %>% 
  group_by(my_weekdays) %>% 
  summarize(n = n()) %>% 
  ggplot() + 
  geom_col(aes(x = my_weekdays, y = n))

crashes_2017 %>% 
  mutate(my_hours = hour(hm(time))) %>% 
  group_by(my_hours) %>% 
  summarize(n = n()) %>% 
  ggplot() + 
  geom_col(aes(x = my_hours, y = n))

crashes_2017 %>% 
  mutate(my_weekdays = weekdays(date), my_hours = hour(hm(time))) %>% 
  group_by(my_weekdays, my_hours) %>% 
  summarise(n = n()) %>% 
  ggplot() + 
  geom_line(aes(x = my_hours, y = n, col = my_weekdays), size = 1.05)
# the legend needs some reordering
```

```{r, echo=FALSE, eval=FALSE}
crashes_2019 %>% 
  select(matches("time|date")) %>% 
  lapply(head, 10)
crashes_2019 %>% 
  select(matches("time|date")) %>% 
  sapply(class)
```
Of the three time variables, it should be clear from the outcome of previous exercises that `datetime` is contains the most useful information.
To consolidate the plotting know-how learned in Chapter \@ref(pkgs), let's start by simply plotting the datetime object, a good way to understand new datasets and the variables they contain.
Create the following three plots to show how `date` and `time` vary as a function of `datetime`:

```{r ggtime1, warning=FALSE, fig.cap="Three plots of the datetime (x axis) in relation to the date and time axis.", out.width="30%", fig.show='hold'}
library(ggplot2)
ggplot(crashes_2019) + geom_point(aes(datetime, date))
ggplot(crashes_2019) + geom_point(aes(datetime, time))
b = c("07:00", "09:00", "12:00", "17:00", "19:00")
ggplot(crashes_2019) + geom_point(aes(datetime, time), alpha = 0.01) +
  scale_y_discrete(breaks = b)
```

Figure \@ref(fig:ggtime1) tells us many things about the contents of the three temporal variables, and even provides insight into the temporal distribution of road casualties in Great Britain.
The first two plots show 1) that the `date` variable is identical to the `datetime` variable (at least on the daily resolution than can be seen on the graph) and 2) that `time` values repeat regularly for the range of dates in datetime (from the start of Jan 2019 to end of Dec 2019).
The third plot, which makes use of `ggplot2`'s functionality to show only certain labels on the Y axis and reduced oppacity so that overlapping points are not completely black, is by far the most useful.
It shows that most crashes happen between around 7am and 7pm, with a 'long tail' of crashes in the evening, and that for most of the year there is a clear weekly cycle, reflecting the uptick in crashes during the rush hour commute on weekdays, a pattern that is greatly diminished during several weeks in summer (perhaps corresponding with summer holidays).
The 52 weeks of the year can be distinguished even in this small and simple plot, highlighting the ability of visualisation to help understand data.
Next, let's look at how the time-of-day that crashes happen varies as a function of season, severity and day of week.

From a datetime object of class `POSIXct` can be extracted any type of time information, including the minute, hour, day of week, and month of the crash (or other) event that the object records.

```{r, eval=FALSE, echo=FALSE}
library(lubridate)
weekdays(min(crashes_2019$date) + 0:6)
table(factor(weekdays(crashes_2019$date), levels = weekdays(min(crashes_2019$date) + 0:6)))
```

Building on the time series plot we created in Section \@ref(ggplot2), let's create a graph showing how the hourly distribution of crash numbers changes during the course of a working week.
We will do this first by *preprocessing* the data, creating a new object called `crashes_dow` containing `hour` and `day` columns, and then plotting the result, after filtering out the weekend days, as shown in the code chunk below which results in Figure \@ref(fig:ggfacetwkday):

```{r ggfacetrwkay, fig.cap="Facetted time series showing how the number of crashes increases during the working week.", warning=FALSE, message=FALSE, fig.height=2}
# days of the week:
dow = c("Sunday", "Monday", "Tuesday", "Wednesday", "Thursday", "Friday", "Saturday")
crashes_dow = crashes_2019 %>% 
  mutate(hour = lubridate::hour(datetime)) %>% 
  mutate(day = factor(weekdays(date), levels = dow)) 

crashes_dow %>% 
  filter(!is.na(hour) & !day %in% c("Saturday", "Sunday")) %>% 
  ggplot(aes(hour)) +
  geom_bar(width = 1.01) +
  facet_wrap(~day, nrow = 1)
```

The result is useful, but if we're interested in the number of crashes per hour on different days of the week *relative to the average*, we need to undertake one more preprocessing step.
We will count the number of crashes per hour for all 5 working days, divided by 5 to get the average, then count the number of crashes per hour/week combination and then divide the latter by the former.
These steps are shown in the code chunk below, which results in Figure \@ref(fig:ggfacetrwkayrel).

```{r ggfacetrwkayrel, fig.cap="Facetted time series showing relative number of crashes per hour by day in the working week.", warning=FALSE, message=FALSE, fig.height=2}
crashes_dow_rel = crashes_dow %>% 
  filter(!is.na(hour) & !day %in% c("Saturday", "Sunday")) %>% 
  select(day, hour) %>% 
  group_by(hour) %>% 
  mutate(n_per_hour = n()/5) %>% 
  group_by(day, hour) %>% 
  summarise(n_hday = n(), n_h = first(n_per_hour)) %>% 
  mutate(hday_relative = n_hday / n_h)
summary(crashes_dow_rel)
crashes_dow_rel %>% 
  ggplot() +
  geom_col(aes(hour, hday_relative)) +
  facet_wrap(~day, nrow = 1)
  
```

The results clearly show that Friday is a dangerous day.
The extent to which the high relative number of crashes in the most anomalous hours, on Friday evening, is due increased exposure vs increased risk per km travelled cannot be ascertained by this plot but it certainly suggests that Friday afternoon and evening is a worthy focus of road safety research.

Exercises: 

1. Building on the code above, show the absolute and relative number of crashes per hour on Saturday and Sunday.
2. Filter the dataset so it contains only data from two police forces of your choice (e.g. `West Yorkshire` and `Metropolitan Police`).
3. Try creating plots similar to those shown above but facetted by police force rather than by day of the week.



## Handling dates and date-times

it is worth remembering that base R already has decent support for dates and `datetimes`, although the base R functions are not particularly intuitive.
This is shown in the code chunk below, which creates objects representing the date and time of a fictitious crash event on a cold winter's morning, 1^st^ January 2020, and a subsequent road safety intervention on the 20^th^ October 2020:

```{r}
crash_datetime_character = "2020-01-01 08:35"
crash_datetime = as.POSIXct(crash_datetime_character)
class(crash_datetime)
intervention_date_character = "2020-10-20"
intervention_date = as.Date(intervention_date_character)
class(intervention_date)
# see ?as.POSIXct for more examples
```

Fortunately for most users, there are easier ways to work with time series data, starting with the `hms` package.

## Hours, minutes seconds with hms

The `hms` library can be used to process hours, minutes and seconds, as shown below.
See a very basic demo of the package and links to the package's help pages with the following commands (in which we use the package without loading it with the `library()` function, demonstrating the `package::function()` syntax taught in Chapter \@ref(pkgs):

```{r}
library(tidyverse)
```

```{r}
crash_time_character = "08:35:00"
crash_time_hms = hms::as_hms(crash_time_character)
class(crash_time_hms)
```

```{r, eval=FALSE}
?hms::`hms-package`
```

As the package's name suggests it is used for dealing with hours minutes and seconds.
It can round time objects of class `hms` to the nearest second (or any multiple of a second):

```{r}
hms::round_hms(crash_time_hms, 1)           # time to the nearest second
hms::round_hms(crash_time_hms, 1 * 60 * 60) # time to the nearest hour
hms::round_hms(crash_time_hms, 1 * 30 * 60) # time to the nearest half hour
```
It can also convert simple text strings into time objects, e.g. as follows (note we do not need to include the `:00`):

```{r}
hms::parse_hm("08:35")
```

## The lubridate package

In many cases the most useful package easy of use when working with temporal data is `lubridate`.
Load it as follows:

```{r, message=FALSE}
library(lubridate)
```

The simplest example of a Date object that we can analyze is just the current date, i.e.

```{r}
today()
```

We can manipulate this object using several `lubridate` functions to extract the current day, month, year, weekday and so on...

```{r, eval=FALSE}
x = today()
day(x)
wday(x)
# Base R function to get the day of week
weekdays(x)
wday(x) %in% c(1, 6) # is it the weekend?
month(x)
year(x)
```

Exercises: 

1. Look at the help page of the function `month` to see how it is possible to extract the current month as character vector 
1. Look at other functions in lubridate to extract the current weekday as a number, the week of year and the day of the year

Date variables are often stored simply as a character vectors.
This is a problem, since R is not always smart enough to distinguish between character vectors representing Dates.
`lubridate` provides functions that can translate a wide range of date encodings such as `ymd()`, which extracts the Year Month and Day from a character string, as demonstrated below.

```{r, eval=FALSE}
as.Date("2019-10-17") # works
as.Date("2019 10 17") # fails
ymd("2019 10 17")     # works
dmy("17/10/2019")     # works
```

Import function such as `read_csv` try to recognize the Date variables.
Sometimes this fails.
You can manually create Date objects, as shown below.

```{r}
x = c("2009-01-01", "2009-02-02", "2009-03-03")
x_date = ymd(x)
x_date
```

Exercises: 

1. Extract the day, the year-day, the month and the weekday (as a non-abbreviated character vector) of each element of `x_date`. 
1. Convert `"09/09/93"` into a date object and extract its weekday. 
1. **Bonus:** Read the help page of `as.Date` and `strptime` for further details on base R functions for dates. 
1. **Bonus:** Read the Chapter 16 of [R for Data Science book](https://r4ds.had.co.nz/dates-and-times.html) for further details on `lubridate` package. 

```{r, echo=FALSE, eval=FALSE}
# 1. Extract the day, the year-day, the month and the weekday (as a non-abbreviated character vector) of each element of `x_date`. 
day(x_date)
yday(x_date)
month(x_date)
weekdays(x_date, abbreviate = FALSE)
# 1. Modify the previous example to parse the following character string: `"09/09/1993"` and extract its weekday. 
weekdays(dmy("09/09/93"))
wday(dmy("09/09/93"))
```

## Dates in a data frame

We can use Dates also for subsetting events in a dataframe. For example, if we define `x_date` as before and add it to the `crash` dataset, i.e.

```{r}
crashes$casualty_day = x_date
```

then we can subset events using Dates. For example

```{r}
filter(crashes, day(casualty_day) < 7) # the events that ocurred in the first week of the month
filter(crashes, weekdays(casualty_day) == "Monday") # the events occurred on monday
```

Exercises: 

1. Select only the events (rows in `crashes`) that occurred in January
1. Select only the events that ocurred in an odd year-day 
1. Select only the events that ocurred in a leap-year (HINT: check the function `leap_year`)
1. Select only the events that ocurred during the weekend or in June
1. Select only the events that ocurred during the weekend and in June
1. Count how many events ocurred during each day of the week. 

## Components of time objects

Now we'll take a look at the time components of a Date. Using the function `hms` (acronym for Hour Minutes Seconds) and its subfunctions such as `hm` or `ms`, we can parse a character vector representing several times as an Hour object (which is tecnically called a Period object). 

```{r}
x = c("18:23:35", "00:00:01", "12:34:56")
x_hour = hms(x)
x_hour
```

We can manipulate these objects using several `lubridate` functions to extract the hour component, the minutes and so on:

```{r}
hour(x_hour)
minute(x_hour)
second(x_hour)
```

If the Hour data do not specify the seconds, then we just have to use a subfunction of `hms`, namely `hm`, and everything works as before. 

```{r}
x = c("18:23", "00:00", "12:34")
(x_hour = hm(x))
```

We can use Hour data also for subsetting events, like we did for Dates. Let's add a new column to crashes data, 

```{r}
crashes$casualty_hms = hms(c("18:23:35", "00:00:01", "12:34:56"))
crashes$casualty_hour = hour(crashes$casualty_hms)
```

Exercises: 

1. Filter only the events that ocurred after midday (i.e. the PM events). Hint: your answer may include `>= 12`.
1. Filter only the events that ocurred between 15:00 and 19:00
<!-- 1. Round all hours to the next hour. Hint: Look at the help page of the `round_date` function.  -->
1. **Bonus (difficult):** run the following code, which downloades data for car crashes occurred during 2017.
